---
title: "Introduction to pandas"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing the `pandas` library

```{r, echo=FALSE}
library(reticulate)

use_condaenv("r-reticulate")
```

When we import `pandas` in Python, we generally use the `pd` acronym. As such, instead of writing `pandas.shape` we can just type `pd.shape`. As you've noticed this is just a matter of convenience nonetheless it's important to follow the community's conventions, that's why I highly encourage you to always work with the `pd` acronym.

```{python}
import pandas as pd
```

# Reading a `csv` file in pandas 

Using the `read_csv()` function we can read either a local `csv` file or as in the following a remote `csv` file. We'll work with the __penguins__ data frame which is provided by the  [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/) `R` package. 

> feel free to copy paste the URL below to check out the data frame in a raw format

```{python}
URL = "https://gist.githubusercontent.com/feddelegrand7/2f32818389d58eb62e8e47ff08751415/raw/9c2debfc2277aba6c227a0b1746f4af78c9798fe/penguins.csv"
penguins = pd.read_csv(URL)
```


# Getting an overview of the data frame 

Before any analysis, it's useful to get some basic information about the structure of a data frame. Here `df.shape()` will print out the number of rows and columns in a `tuple` format: 


```{python}
penguins.shape
```
OK ! it's obvious that our `penguins` data frame contains 344 rows and 8 columns. Now, will use `penguins.head(3)` in order to display the first 3 rows: 

```{python}
penguins.head(3)
```

If you're working with a data frame that is too wide and you need a list of the columns' names, you can use the `columns` attribute: 


```{python}
penguins.columns
```

It's also possible to print out the row-index of the data frame using the `index` attribute. 


```{python}
penguins.index
```


The `info()` method is a pretty cool tool that helps get an overall understanding of the structure of our data. It prints out in order: 

- The class of the data frame : `DataFrame` 
- The range of the index: `344` rows
- The number of __non-missing__ values, here spotted as `non-null` 
- The type of each column or `Dtype`. Here we have two types of columns: 
  1. `object`: you can think of it as a character string column; 
  2. `float64`: a numeric column
- Lastly, we get the `memory usage` which provides an idea about the memory size of our data frame. 


```{python}
penguins.info()

```
 


# Selecting a column in pandas 

There are two ways of selecting columns in `pandas`, we can do: 

1. `df['column_name']` or: 
2. `df.column_name`.

for many reasons that will become obvious later, you should always use the first syntax.


In our example, let's say that we wanted to select the `species` column, we can do: 

```{python}

penguins['species']

```

__Important__ `r emo::ji('alembic')` : a column taken individually has a special form and a special type called `pandas Series`. To summarize a `pandas` data frame has a `DataFrame` class and is composed of several columns that taken __individually__ have a `Series` class. 

```{python}
type(penguins['species'])
```


Now let's select multiple columns, maybe `species`, `island` and `bill_length_mm`.


```{python, error=TRUE}
penguins['species', 'island', 'bill_length_mm']

```

We got an error `r emo::ji('angry')` !!! what happened ? in fact there is a dialogue here: 

> pandas: you want to select multiple column ? 
> me: Yes! please !!!
> pandas: OK! but I want a __list []__ of columns, in other words, the columns must be provided inside a list ! otherwise you'll get an error !!!!
> me: emo::ji('unamused')


So the correct way of subsetting columns is: 

```{python}
penguins[['species', 'island', 'bill_length_mm']]

```

Works like a charm `r emo::ji('unamused')`

# Removing a column 


We've a perfect data frame to play, why would we dare remove a column or two ??? just for the sake of the example ... ok that's was the joke of the day, now let's get into serious stuffs. 

The `drop()` method allows us to remove the columns that we specify. Let's get rid of `species` and `island`: 

```{python}
penguins_drop = penguins.drop(columns= ['species', 'island'])
penguins_drop.head(5)
```


# Renaming columns 

When renaming a column, `pandas` expect that you define a `dictionary` as follows: 

`{'current_name' : 'new_name'}`

So if for example, we wanted to change the name of `species` and  `island` to respectively `penguin_name` and `land`, we can proceed as follows: 


```{python, message = FALSE, message=FALSE, warning=FALSE}
penguins.rename({'species': 'penguin_name', 'island':'land'}, 
                axis = 'columns')
```
and voila ! here we had to set `axis = 'columns'` to specify that we're renaming columns not indexes (which is the default)